\section{Project description ($\pm$ 10\% total words) }
\subsection{Domain}
The report draws on a variety of disciplines. However, it is impossible to appreciate the vastness and significance of the topic in a limited number of pages. Nevertheless, we attempt to make the report self-contained by providing the reader with an intuitive understanding of our survey and application results.
The domains that are associated with our project are: 
\begin{itemize}
\item Internet security: a branch to computer security, related to the Internet, comprised of measures to protect transactions done over a collection of interconnected networks \cite{is}.
\end {itemize}
\begin{itemize} 
\item Computer Security: the process of preventing and detecting unauthorized use of a computer \cite{cs}. 
\end{itemize}
\begin{itemize}
\item Cryptographic primitives:  well-established, low-level cryptographic algorithms frequently used to build cryptographic protocols for computer security systems \cite{cp}.
\end{itemize}
\begin{itemize}
\item Symmetric-key cryptography: an algorithm for cryptography that uses the same key for both encryption and decryption \cite{ska}.
\end{itemize}
\begin{itemize}
\item Cryptographic algorithms: ensure alteration of data in order to preserve its authenticity, confidentiality, integrity through various security mechanisms such as symmetric encryption,  asymmetric encryption and cryptographic hash functions \cite{wikicrypto}.
\end{itemize}




\subsection{Objectives}
The objectives of this project are three-fold. 
\begin{enumerate}
\item  As a preliminary task, we address a highly needed background material. We give a basic introduction to the concepts and primitives of cryptography, i.e. transposition/permutation, encryption/decryption, symmetry/asymmetry, block ciphers/stream ciphers and how they preserve their security in constantly emerging attacks.
\item  Our main focus falls on symmetric encryption, taking into account both classical and modern algorithms. Therefore,  we review existing state-of-the-art encryption algorithms: Caesar cipher, ROT13, One-time pad, DES, 3DES, AES.
\item Last but not least, we implement a chat system that encrypts user's messages before they are transmitted over the Internet.
\end{enumerate}


\subsection{Constraints}
Our project takes into account substitution/transposition; block/stream ciphers but we do not consider:
\begin{itemize}
\item Asymmetric ciphers (public-key algorithms, including RSA and elliptic curve); 
\end{itemize}
\begin{itemize}
\item Data integrity algorithms (cryptographic hash functions, message authentication codes (MACs); digital signatures; 
\end{itemize}
\begin{itemize}
\item Authentication techniques: key management and key distribution (protocols for key exchange);
\end{itemize}


